---
title: "Unit 2-Real Data Cleaning ii"
author: "Alan Thomas Joshy"
date: "2026-04-08"
output: html_document
knit: (function(input, ...) rmarkdown::render(input, output_dir = "../outputs"))
---

This document applies the Unit 2 data cleaning workflow to a second real AIMS dataset — the Long Term Monitoring Program (LTMP) benthic cover dataset. Unlike the previous cleaning exercise, this dataset presented a different structural challenge: the data arrives in long format, with four separate rows per site visit representing different benthic cover categories. The cleaning workflow here therefore includes reshaping the data to wide format as a core step, in addition to the standard inspection, renaming, and type correction steps.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
```



```{r}
cover <- read_csv("../data/ltmp_hc_sc_a_by_site/ltmp_hc_sc_a_by_site.csv")

head(cover)

str(cover)

summary(cover)
```

Initial inspection shows 18,540 rows across 12 columns spanning surveys from 1992 to 2023. read_csv() has correctly detected most column types, including SAMPLE_DATE as a date. However SITE_NO and YEAR_CODE are read as numeric despite being identifiers rather than measurements — this will need correcting. The SHELF and SECTOR columns contain short codes that will benefit from recoding to more meaningful labels.


```{r}
#Checking for duplicate rows 
sum(duplicated(cover))
```

Before renaming or recoding anything, it is worth checking for duplicate rows — a common issue in large monitoring datasets where records may have been entered twice. sum(duplicated()) returns 0, confirming no duplicates are present.

The janitor package's clean_names() function converts all column names to snake_case in a single step — a useful shortcut compared to manually renaming each column with rename(). This is particularly helpful for wide datasets with many columns.

```{r}
#Renaming column names to snake_case
cover <- cover |> clean_names()
names(cover)

#Year code and Site no are not actually numerical values or as they are more of a label or identifier than a measurement so we change the column types to character
cover <- cover |> mutate(
  year_code = as.character(year_code), site_no = as.character(site_no)
)

str(cover)
```

SITE_NO and YEAR_CODE are converted from numeric to character using as.character() inside mutate(). Although they contain numbers, neither represents a measurement — SITE_NO is a transect identifier and YEAR_CODE is a survey period label (e.g. 199293 represents the 1992-93 survey year). Treating them as numbers would be misleading and could cause issues in analysis.

```{r}
#Running unique() on the character columns to check
unique(cover$sector)
unique(cover$shelf)
unique(cover$group_code)
unique(cover$site_no)

cover <- cover |> 
  mutate(shelf = case_when(
    shelf == "I" ~ "Inshore",
    shelf == "M" ~ "Midshelf",
    shelf == "O" ~ "Offshore"
  ))
```


Checking unique values in categorical columns confirms consistent coding — no spelling errors or duplicate categories. The group_code column contains four categories: Algae, Hard Coral, Soft Coral, and Other. The shelf column contains three codes — "I", "M", and "O" — representing Inshore, Midshelf, and Offshore positions across the GBR shelf gradient. This is an ecologically important variable as reef condition, coral cover, and disturbance history vary considerably across the shelf. These codes are recoded to full descriptive labels using case_when() for clarity.
The sector column contains ten codes representing geographic sectors along the GBR. Without confirmed metadata these are left as-is, though based on standard AIMS sector designations they likely correspond to regions including Cairns, Townsville, Whitsundays, and Swains among others.

```{r}
cover_wide <- cover |> 
  pivot_wider(names_from = group_code,
              values_from = cover)

cover_wide <- cover_wide |> clean_names()
names(cover_wide)
```

The core structural issue with this dataset is its long format — each site visit occupies four rows, one per benthic cover category. While long format is useful for certain analyses and plotting, wide format is more intuitive for inspection and many analytical workflows. pivot_wider() reshapes the data so each site visit becomes a single row with separate columns for algae, hard coral, soft coral, and other cover. This reduces the dataset from 18,540 rows to 4,635 rows — exactly one quarter, confirming the pivot worked correctly. clean_names() is applied again to tidy the new column names generated by the pivot.

```{r}
colSums(is.na(cover_wide))
summary(cover_wide)
nrow(cover_wide)

cover_wide |> filter(visit_no == 0)
```

Final inspection confirms no missing values across any column — a notably clean result for a real world dataset of this size. Cover values are all within ecologically realistic ranges for GBR reef surveys. One anomaly worth noting is that 15 rows across 5 reefs have visit_no == 0, all from June 1992 — the earliest surveys in the dataset. This likely reflects a coding convention used at the start of the monitoring program before the visit numbering system was formalised, rather than a data error. These rows contain valid cover measurements and are retained, but flagged here for transparency.