---
title: "R Learning Project-Unit 2"
author: "Alan Thomas Joshy"
date: "2026-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


##Tidyverse

```{r}
reef <- data.frame(
  site = c("Pelorus", "Orpheus", "Magnetic", "Dunk", "Heron",
           "Pelorus", "Orpheus", "Magnetic", "Dunk", "Heron"),
  year = c(2022, 2022, 2022, 2022, 2022,
           2023, 2023, 2023, 2023, 2023),
  coral_cover = c(42, 35, 18, 55, 60,
                  38, 33, 15, 58, 62),
  fish_abundance = c(120, 98, 45, 210, 185,
                     110, 95, 40, 220, 190),
  mean_temp = c(26.3, 27.1, 25.8, 27.6, 26.9,
                26.8, 27.4, 26.1, 28.0, 27.2)
)
```


Using filter() - keeps rows matching a conditon
```{r}
# Show only surveys from 2022
reef |>
  filter(year == 2022)

# Show only sites with coral cover greater than 40%
reef |> 
  filter(coral_cover > 40)
```

```{r}
# Coral cover greater than 40% AND (&) the year is 2023
reef |> 
  filter(coral_cover > 40 & year == 2023)

# Mean temp greater than 27C OR (|) fish abundance greater than 200
reef |> 
  filter(mean_temp > 27.0 | fish_abundance > 200 )

# & and |, used in filter

```


Using select() - chooses columns to keep
```{r}
# select only site and coral cover columns
reef |> 
  select(site, coral_cover)

# select everything except fish abundance ( use ,-, infront of a column to exclude it)
reef |> 
  select(-fish_abundance)
```


Using mutate() - adds a new column or modifies an existing one. very useful in environmental work for calculating derived variables.
```{r}
# Add a new column called coral_change that calculates the difference between coral_cover and a baseline of 50%. This would tell you how far above or below a target each site is
reef |> 
  mutate(coral_change = coral_cover - 50)
# Add a new column converting temperature from Celsius to Fahrenheit
reef |> 
  mutate(mean_F = (mean_temp * 9/5) + 32)
# OR we can update a column instead of making a new one
reef |> 
  mutate(mean_temp = (mean_temp * 9/5) + 32)
```


Using arrange() - simply sorts the dataframe by a column
```{r}
# Arrange by coral cover from lowest to highest
reef |> 
  arrange(coral_cover)

#Arrange by Highest to lowest
reef |> 
  arrange(desc(coral_cover))
```


Using summarise() - collapses the entire data frame down to a single summary row. You specify what summary stats you want
```{r}
# Calculate the mean and maximum coral cover across all sites and years
reef |> 
  summarise(mean_coral = mean(coral_cover),
            max_coral = max(coral_cover))

# Calculate the mean and maximum fish abundance across all sites and years
reef |> 
  summarise(mean_abundance = mean(fish_abundance),
            max_abundance = max(fish_abundance))

# Using filter() with summarise() can allow us to select certain variables and calculate summary stats
reef |> 
  filter(year == 2022) |> 
  summarise(mean_abundance_2022 = mean(fish_abundance),
            max_abundance_2022 = max(fish_abundance))

#mean coral coer in 2023 
reef |> 
  filter(year == 2023) |> 
  summarise(mean_cover_2023 = mean(coral_cover))
```


group_by() is used mostly with other verbs like summarise(), mutate(), or filter().
```{r}
# Calculate the mean coral cover of each site
reef |> 
  group_by(site) |> 
  summarise(mean_coral = mean(coral_cover))

# Calculate how much the the coral cover deviates from the mean coral cover for each site
reef |> 
  group_by(site) |> 
  mutate(coral_cover_mean_difference = coral_cover - mean(coral_cover))
```

One important thing to remember — after using group_by() your data frame stays grouped until you explicitly ungroup it with ungroup(). This can cause unexpected behaviour down the line if you forget, so it's good practice to add |> ungroup() at the end when you're done:
```{r}
reef |> 
  group_by(site) |> 
  summarise(mean_coral = mean(coral_cover)) |> 
  ungroup()
```

